---
title: "Using Shiny to create apps and dashboards"
author: "Václav Čepelák"
date: "April 9, 2018"
runtime: shiny
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Outline

- What is Shiny?
- On what platforms it can be used?
- How to build the app?
- How to improve graphics?

# What is Shiny?

## Shiny

Shiny is an R package that builds interactive webapps straight from R

```{r install, echo = TRUE, eval = FALSE}
install.packages('shiny')
library(shiny)
```

[https://shiny.rstudio.com](https://shiny.rstudio.com)
gallery of Shiny apps with source codes [here](https://shiny.rstudio.com/gallery/)
Shiny tutorials [here](https://shiny.rstudio.com/tutorial/)


## What platforms?

- Run directly from R
- shinyapps.io
- RStudio connect
- install it on a server
- [embed](https://beta.rstudioconnect.com/content/2671/Combining-Shiny-R-Markdown.html) in R Markdown (html, ioslides, [flexdashboard](https://rmarkdown.rstudio.com/flexdashboard/))


## How it works

```{r basic, echo = TRUE, eval = FALSE}
library(shiny)

ui <- fluidPage(
   # Define UI for application: inputs and outputs
)

server <- function(input, output) {
  # Define what computations need to be performed
  # i.e. what happens if you change inputs
}

# Run the application 
shinyApp(ui = ui, server = server)
```


## ... and for more complex apps

The app is composed from multiple `.R` files and additional files required for the app (e.g. data, pictures, CSS) stored in one folder having the name of the app.

- `global.R` is an optional script, which performs all the computations before the app is launched
- `ui.R` is a script that defines the user interface
- `server.R` is a script that defines the computations performed
- `www` is a folder containing all the supplementary files

The app can be run by clicking on the `Run app` button or by
```{r runapp, echo = TRUE, eval = FALSE}
runApp('path/to/folder')
```



## Let's build a simple app together

We are going to build a simple app: takes an SPSS data file as an input and plots a frequency of a selected variable.

We will follow these steps:
1. Decide how the UI is going to look like
2. Prepare the script to load data
3. Set the functionalities of the app to run properly
4. Run the app


# User interface


## Building a user interface

1. **General layout** (sidebars, panels, pages etc.)
2. **Widgets** (sliders, input boxes etc.) 
3. **Outputs** (plots, tables etc.)

## General layout

Shiny is HTML-based, i.e. R functions are used to build an HTML code. Therefore anyone familiar with HTML can customize Shiny apps easily (R contains mirror-functions to HTML tags). 

There are built-in functions to compose layouts:
- `sidebarLayout` (a simple one - **let's use it for our app**)
- grid layout (using `fluidRow` and `columns` function)

Multiple panels can be used:
- using `tabsetPanel` function
- using `navlistPanel`

more information [here](https://shiny.rstudio.com/articles/layout-guide.html)


## Our app's layout will look like this

```{r layout_structure, echo = TRUE, eval = FALSE}

ui <- fluidPage(
  # Title of the app
  titlePanel("My first app"), 
  
  # Define the sidebar layout
  sidebarLayout(
    
    sidebarPanel(
      # add content (the widgets) here
    ),
    
    mainPanel(
     # add content (the outputs) here 
    )
  )
)

```


## Widgets

The gallery of widgets available [here](https://shiny.rstudio.com/gallery/widget-gallery.html)

A widget is created by a function (e.g. a select box is defined by the [`selectInput` function](https://shiny.rstudio.com/reference/shiny/1.0.3/selectInput.html))

Each object (widget, plot) has its unique name (id) defined in the function


## Our app's widgets

We will add a select box that selects a variable from input data and a radio button with ordering options

```{r input_widgets, echo = TRUE, eval = FALSE}

    sidebarPanel(
      # add select box
      selectInput('variable', label = 'Select variable: ',
                  choices = ## will be added later ##
      ),
      # add radio button
      radioButtons('order', label = 'Ordering options: ',
                   choices = c('None', 'Sort by frequency', 
                               'Sort alphabetically'),
                   selected = 1)
      
    ),

```


## Two remarks on widgets

Sliders in Shiny (`sliderInput`) have a `playButton` (and a `pauseButton`) parameter which enables to do a simple animated plot (more details [here](https://shiny.rstudio.com/gallery/sliders.html))

*File upload* and *file download* are possible in the Shiny app (examples [here](https://shiny.rstudio.com/gallery/file-upload.html) and  [here](https://shiny.rstudio.com/gallery/file-download.html))


## Add logo

We will add a short text and logo using the `img` and `h4` functions.
These functions create the `<img>` and `<h4>` HTML tags and the argument passed to them are exactly the same as would be passed to [these tags](https://www.w3schools.com/tags/att_img_width.asp).

```{r picture, echo = TRUE, eval = FALSE}
    sidebarPanel(
      ### Widgets ###
      # name of a file stored in the `myfirstapp/www` folder
      img(src = 'Ipsos-logo.jpg', align = "center", height = "42" width = "42"),
      h4('Created by ', b('Ipsos'))
    ),
```




## Outputs {.flexbox .vcenter}

Outputs are displayed in the app using `*Output` functions.
Each `*Output` function has its mirror `render*` function being used in the `server.R` script

| **Type of content** |      **Output**     |  **Render**  |
|:-------------------:|:-------------------:|:------------:|
|   plot (graphical)  |      plotOutput     |  renderPlot  |
|        plotly       |     plotlyOutput    | renderPlotly |
|        image        |     imageOutput     |  renderImage |
|        table        |     tableOutput     |  renderTable |
|      data table     |   dataTableOutput   |              |
|         HTML        | htmlOutput uiOutput |   renderUI   |
|       R output      |  verbatimTextOutput |  renderPrint |
|         Text        |      textOutput     |  renderText  |


## Plotly

[https://plot.ly](https://plot.ly)
- An open source platform for creating interactive plots that can be embedded in HTML documents
- Libraries for R, Python, JavaScript
- A Plotly tutorial for Shiny [here](https://plot.ly/r/shiny-tutorial/)

## a Plotly example

```{r plotly_example, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}
library(plotly)
p <- plot_ly(data = iris, x = ~Sepal.Length, y = ~Petal.Length,
             marker = list(size = 10,
                           color = 'rgba(255, 182, 193, .9)',
                           line = list(color = 'rgba(152, 0, 0, .8)',
                                       width = 2))) %>%
  layout(title = 'Styled Scatter',
         yaxis = list(zeroline = FALSE),
         xaxis = list(zeroline = FALSE))
p

```


## Our app's outputs

Let's create a main panel. We will add a header containing the question's label - this is performed via the `uiOutput` function. And we will add a `plotly` bar chart.

```{r main_panel, echo = TRUE, eval = FALSE}
mainPanel(
     uiOutput('question'),
     plotlyOutput('plot')
    )
```


# Loading the data

## Loading the data

The data are usually loaded and transformed in the `global.R` script (more on this later). 

The data are loaded using the R functions from any format R is able to read:
- `read.csv` / `read.csv2` for **csv** files
- `read.table` for **txt** files
- `read.spss` from the `foreign` library for **sav** files

## Our app's data

The following script loads data from SPSS to R and selects only factor variables with max. 10 distinctive categories.

```{r main_panel1, echo = TRUE, eval = FALSE}
### global.R script ###
library(foreign)
shiny_data <- read.spss('./www/wvs1995_v3.sav', to.data.frame = TRUE)
id_var <- 'V3'

selected_columns <- sapply(names(shiny_data), function(x) {
  if (x == id_var) {
    return(TRUE)
  } else if (attr(shiny_data, 'variable.labels')[x]) == '') {
    return(FALSE)
  } else if (!is.factor(shiny_data[[x]])) {
    return(FALSE)
  } else {
    return(length(levels(shiny_data[[x]])) <= 10)
  }
})
### =>
```

## Our app's data

```{r main_panel2, echo = TRUE, eval = FALSE}
### =>
variables <- data.frame(var = names(shiny_data)[selected_columns],
                        label = attr(shiny_data,
                                     'variable.labels')[selected_columns],
                        stringsAsFactors = FALSE)

shiny_data <- shiny_data[, selected_columns]


```

```{r selectbox, echo = TRUE, eval = FALSE}
# choices are added to the selectInput
selectInput('variable', label = 'Select variable: ',
            choices = variables$var),

```


# Server function

## How the app works

- The `server` function (re)computes outputs when inputs change
- Code outside the `server` functions runs **once per R session** (... it means that when underlying data change the app must be restarted)
- Code inside the `server` function runs **everytime one user connects**
- Code inside a reactive function (i.e. function inside the `server` function) runs **with every reaction** (change of the input)


## Render functions

The render functions are placed inside the server function like this...

```{r server_function, echo = TRUE, eval = FALSE}

server <- function(input, output) {
  output$question <- renderUI({
    ### code here ###
  })
  
  output$plot <- renderPlotly({
    ### code here ###
  })
}

```


## The renderUI function

This code creates the header with the varable label...

```{r renderui, echo = TRUE, eval = FALSE}

output$question <- renderUI({
  ui_output <- h2(variables$label[variables$var == input$variable])
  return(ui_output)
})

```



## Reactive function

The `reactive` function is used to create objects (transformations, tables, subsets of data etc.) **inside** the `server` **function**, which is recomputed with the change of inputs but does not directly print any output but is used by other `render` functions.

Note: the `reactive` objects are called in other parts of code as **functions**

More details [here](https://shiny.rstudio.com/tutorial/written-tutorial/lesson6/).

## Reactive function in our app

The reactive function is used for preparing the underlying table of the plot

```{r reactive, echo = TRUE, eval = FALSE}

plot_table <- reactive({
  plot_table <- data.frame(table(shiny_data[[input$variable]]))
  if (input$order == '2') {
    plot_table <- dplyr::arrange(plot_table, desc(Freq))
  } else if (input$order == '3') {
    plot_table <- dplyr::arrange(plot_table, Var1)
  }
  return(plot_table)
})

```




## The renderPlotly function

This code creates the plot...

```{r renderplotly, echo = TRUE, eval = FALSE}

output$plot <- renderPlotly({
  p <- plot_ly(x = plot_table()$Var1,
               y = plot_table()$Freq,
               type = 'bar',
               orientation = 'h')
})

```


## And the app will look like this

```{r app, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
library(dplyr)
library(foreign)
library(plotly)
library(shiny)

shiny_data <- read.spss('./www/wvs1995_v3.sav', to.data.frame = TRUE)
id_var <- 'V3'

selected_columns <- sapply(names(shiny_data), function(x) {
  if (x == id_var) {
    return(TRUE)
  } else if (attr(shiny_data, 'variable.labels')[x] == '') {
    return(FALSE)
  } else if (!is.factor(shiny_data[[x]])) {
    return(FALSE)
  } else {
    return(length(levels(shiny_data[[x]])) <= 10)
  }
})

variables <- data.frame(var = names(shiny_data)[selected_columns],
                        label = attr(shiny_data,
                                     'variable.labels')[selected_columns],
                        stringsAsFactors = FALSE)
shiny_data <- shiny_data[, selected_columns]


ui <- fluidPage(
  # Title of the app
  titlePanel("My first app"), 
  
  # Define the sidebar layout
  sidebarLayout(
    
    # add select box
      selectInput('variable', label = 'Select variable: ',
                  choices = variables$var),
      # add radio button
      radioButtons('order', label = 'Ordering options: ',
                   choices = c('None', 'Sort by frequency', 
                               'Sort alphabetically'),
                   selected = 1)
  )
)

server <- function(input, output) {
  
  plot_table <- reactive({
    plot_table <- data.frame(table(shiny_data[[input$variable]]))
    if (input$order == '2') {
      plot_table <- dplyr::arrange(plot_table, desc(Freq))
    } else if (input$order == '3') {
      plot_table <- dplyr::arrange(plot_table, Var1)
    }
    return(plot_table)
  })
  
  output$question <- renderUI({
    ui_output <- h2(variables$label[variables$var == input$variable])
    return(ui_output)
  })
  
  output$plot <- renderPlotly({
  p <- plot_ly(x = plot_table()$Var1,
               y = plot_table()$Freq,
               type = 'bar',
               orientation = 'h',
               width = 500)
  })
}

# Run the app 
shinyApp(ui = ui, server = server)

```
