
---
title: "Using Shiny to create apps and dashboards"
author: "Václav Čepelák"
date: "April 9, 2018"
output: 
  ioslides_presentation:
    widescreen: true
    incremental: true
    css: presentation.css
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# What is Shiny?

## Shiny {.bigfont .build}

Shiny is an R package that builds interactive webapps directly from R

```{r install, echo = TRUE, eval = FALSE}
install.packages('shiny')
library(shiny)
```

<br><h2>[https://shiny.rstudio.com](https://shiny.rstudio.com)</h2>
gallery of Shiny apps with source codes [here](https://shiny.rstudio.com/gallery/)<br>
Shiny tutorials [here](https://shiny.rstudio.com/tutorial/)


## What platforms? {.bigfont}

- Run directly from R
- [shinyapps.io](http://www.shinyapps.io/)
- [RStudio connect](https://www.rstudio.com/products/connect/)
- install it on own server
- [embed](https://beta.rstudioconnect.com/content/2671/Combining-Shiny-R-Markdown.html) in R Markdown (html, ioslides, [flexdashboard](https://rmarkdown.rstudio.com/flexdashboard/))


## How it works

```{r basic, echo = TRUE, eval = FALSE}
library(shiny)

ui <- fluidPage(
   # Define UI for application: inputs and outputs
)

server <- function(input, output) {
  # Define what computations need to be performed
  # i.e. what happens if you change inputs
}

# Run the application 
shinyApp(ui = ui, server = server)
```


## ... and for more complex apps {.build}

The app is composed from multiple `.R` files and additional files required for the app (e.g. data, pictures, CSS) stored in one folder having the name of the app.

- `global.R` is an optional script, which performs all the computations before the app is launched
- `ui.R` is a script that defines the user interface
- `server.R` is a script that defines the computations performed
- `www` is a folder containing all the supplementary files

The app can be run by clicking on the `Run app` button or by
```{r runapp, echo = TRUE, eval = FALSE}
runApp('path/to/folder')
```



## Let's build a simple app together {.build}

We are going to build a simple app: takes an SPSS data file as an input and plots a frequency of a selected variable.

<h4>We will follow these steps:</h4>
<ol>
  <li>Decide how the UI is going to look like</li>
  <li>Prepare the script to load data</li>
  <li>Set the functionalities of the app to run properly</li>
  <li>Run the app</li>
</ol>


# User interface


## Building a user interface {.bigfont}

1. **General layout** (sidebars, panels, pages etc.)
2. **Widgets** (sliders, input boxes etc.) 
3. **Outputs** (plots, tables etc.)

## General layout {.build}

<div>Shiny is **HTML-based**, i.e. R functions are used to build an HTML code. Therefore anyone familiar with HTML can customize Shiny apps easily (R contains **mirror-functions** to HTML tags, more details [here](https://shiny.rstudio.com/articles/tag-glossary.html)). </div>

<div>There are built-in functions to compose layouts:<br>
<ul>
  <li>`sidebarLayout` (a simple one - **let's use it for our app**)</li>
  <li>grid layout (using `fluidRow` and `columns` function)</li>
</ul>
</div>

<div>Multiple panels can be used:<br>
<ul>
  <li>using `tabsetPanel` function</li>
  <li>using `navlistPanel`</li>
</ul>
</div>

<div>more information [here](https://shiny.rstudio.com/articles/layout-guide.html)</div>


## Our app's layout will look like this

```{r layout_structure, echo = TRUE, eval = FALSE}

ui <- fluidPage(
  # Title of the app
  titlePanel("My first app"), 
  
  # Define the sidebar layout
  sidebarLayout(
    
    sidebarPanel(
      # add content (the widgets) here
    ),
    
    mainPanel(
     # add content (the outputs) here 
    )
  )
)

```


## Widgets

- The gallery of widgets available [here](https://shiny.rstudio.com/gallery/widget-gallery.html)

- A widget is created by a function (e.g. a select box is defined by the [`selectInput` function](https://shiny.rstudio.com/reference/shiny/1.0.3/selectInput.html))

- Each object (widget, plot) has its unique name (id) defined in the function

- Two remarks on widgets
    + Sliders in Shiny (`sliderInput`) have a `playButton` (and a `pauseButton`) parameter which enables to do a simple animated plot (more details [here](https://shiny.rstudio.com/gallery/sliders.html))
    + *File upload* and *file download* are possible in the Shiny app (examples [here](https://shiny.rstudio.com/gallery/file-upload.html) and  [here](https://shiny.rstudio.com/gallery/file-download.html))

## Our app's widgets

We will add a select box that selects a variable from input data and a radio button with ordering options

```{r input_widgets, echo = TRUE, eval = FALSE}

    sidebarPanel(
      # add select box
      selectInput('variable', label = 'Select variable: ',
                  choices = ## will be added later ##
      ),
      # add radio button
      radioButtons('order', label = 'Ordering options: ',
                   choiceNames = c('None', 'Sort by frequency', 
                                   'Sort alphabetically'),
                   choiceValues = c('none', 'freq', 'alpha'),
                   selected = 'none')
      
    ),

```


## Add logo

We will add a short text and logo using the `img` and `h4` functions.
These functions create the `<img>` and `<h4>` HTML tags and the argument passed to them are exactly the same as would be passed to [these tags](https://www.w3schools.com/tags/att_img_width.asp).

```{r picture, echo = TRUE, eval = FALSE}
    sidebarPanel(
      ### Widgets ###
      # name of a file stored in the `myfirstapp/www` folder
      tags$img(src = 'Ipsos-logo.jpg', align = "center", height = "42", 
               width = "42"),
      tags$h4('Created by ', b('Ipsos'))
    ),
```




## Outputs

Outputs are displayed in the app using `*Output` functions.
Each `*Output` function has its mirror `render*` function being used in the `server.R` script

```{r table, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}
library(knitr)
df <- data.frame(`Type of content` = c('plot (graphical)', 'plotly', 'image',
                                       'table', 'data table', 'HTML', 
                                       'R output', 'text'),
                 `Output` = c('plotOutput', 'plotlyOutput', 'imageOutput',
                              'tableOutput', 'dataTableOutput', 
                              'htmlOutput uiOutput', 'verbatimTextOutput',
                              'textOutput'),
                 `Render` = c('renderPlot', 'renderPlotly', 'renderImage',
                              'renderTable', 'renderDataTable', 'renderUI', 
                              'renderPrint', 'renderText'))

kableExtra::kable_styling(kable(df, format = 'html'))
```

| **Type of content** |      **Output**     |    **Render**   |
|:-------------------:|:-------------------:|:---------------:|
|   plot (graphical)  |      plotOutput     |   renderPlot    |
|        plotly       |     plotlyOutput    |  renderPlotly   |
|        image        |     imageOutput     |   renderImage   |
|        table        |     tableOutput     |   renderTable   |
|      data table     |   dataTableOutput   | renderDataTable |
|         HTML        | htmlOutput uiOutput |    renderUI     |
|       R output      |  verbatimTextOutput |   renderPrint   |
|         Text        |      textOutput     |   renderText    |


## Plotly {.bigfont}

- [https://plot.ly](https://plot.ly)
- An open source platform for creating interactive plots that can be embedded in HTML documents
- Libraries for R, Python, JavaScript
- A Plotly tutorial for Shiny [here](https://plot.ly/r/shiny-tutorial/)

## a Plotly example

<div align = 'center'>
```{r plotly_example, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}
library(plotly)
p <- plot_ly(data = iris, x = ~Sepal.Length, y = ~Petal.Length,
             marker = list(size = 10,
                           color = 'rgba(255, 182, 193, .9)',
                           line = list(color = 'rgba(152, 0, 0, .8)',
                                       width = 2))) %>%
  layout(title = 'Styled Scatter',
         yaxis = list(zeroline = FALSE),
         xaxis = list(zeroline = FALSE))
p

```
</div>

## Our app's outputs {.bigfont}

Let's create a main panel. We will add a header containing the question's label - this is performed via the `uiOutput` function. And we will add a `plotly` bar chart.

```{r main_panel, echo = TRUE, eval = FALSE}
mainPanel(
     uiOutput('question'),
     plotlyOutput('plot')
    )
```


# Loading the data

## Loading the data

The data are usually loaded and transformed in the `global.R` script (more on this later). 

The data are loaded using the R functions from any format R is able to read:
<ul>
  <li>**read.csv** / **read.csv2** for **csv** files</li>
  <li>**read.table** for **txt** files</li>
  <li>**read.spss** from the **foreign** library for **sav** files</li>
  <li>and many many others</li>
</ul>

## Our app's data

The following script loads data from SPSS to R and selects only factor variables with max. 10 distinctive categories.

```{r main_panel1, echo = TRUE, eval = FALSE}
### global.R script ###
library(foreign)
shiny_data <- read.spss('./www/wvs1995_CZ_subset.sav', to.data.frame = TRUE,
                        use.missings = TRUE)
selected_columns <- sapply(names(shiny_data), function(x) {
  if (attr(shiny_data, 'variable.labels')[x] == '') {
    return(FALSE)
  } else if (!is.factor(shiny_data[[x]])) {
    return(FALSE)
  } else {
    return(length(levels(shiny_data[[x]])) <= 10 & 
             length(levels(shiny_data[[x]])) >= 2)
  }
})
```

## Our app's data

```{r main_panel2, echo = TRUE, eval = FALSE}
### =>
variables <- data.frame(var = names(shiny_data)[selected_columns],
                        label = attr(shiny_data,
                                     'variable.labels')[selected_columns],
                        stringsAsFactors = FALSE)

shiny_data <- shiny_data[, selected_columns]


```

```{r selectbox, echo = TRUE, eval = FALSE}
# choices are added to the selectInput
selectInput('variable', label = 'Select variable: ',
            choices = variables$var),

```


# Server function

## How the app works {.bigfont}

- The `server` function (re)computes outputs when inputs change
- Code outside the `server` functions runs **once per R session** (... it means that when underlying data change the app must be restarted)
- Code inside the `server` function runs **everytime one user connects**
- Code inside a reactive function (i.e. function inside the `server` function) runs **with every reaction** (change of the input)


## Render functions {.bigfont}

The render functions are placed inside the server function like this...

```{r server_function, echo = TRUE, eval = FALSE}

server <- function(input, output) {
  output$question <- renderUI({
    ### code here ###
  })
  
  output$plot <- renderPlotly({
    ### code here ###
  })
}

```


## The renderUI function {.bigfont}

This code creates the header with the varable label...

```{r renderui, echo = TRUE, eval = FALSE}

output$question <- renderUI({
  ui_output <- tags$h2(variables$label[variables$var == input$variable])
  return(ui_output)
})

```



## Reactive function {.bigfont .build}

The `reactive` function is used to create objects (transformations, tables, subsets of data etc.) **inside** the `server` **function**, which is recomputed with the change of inputs but does not directly print any output but is used by other `render` functions.

Note: the `reactive` objects are called in other parts of code as **functions**

More details [here](https://shiny.rstudio.com/tutorial/written-tutorial/lesson6/).

## Reactive function in our app

The reactive function is used for preparing the underlying table of the plot

```{r reactive, echo = TRUE, eval = FALSE}

plot_table <- reactive({
    plot_table <- data.frame(table(shiny_data[[input$variable]]),
                             stringsAsFactors = FALSE)
    if (input$order == 'freq') {
      plot_table <- plot_table %>% dplyr::arrange(desc(Freq))
    } else if (input$order == 'alpha') {
      plot_table <- plot_table %>% dplyr::arrange(as.character(Var1))
    }
    plot_table$Var1 <- factor(plot_table$Var1, 
                              levels = rev(plot_table$Var1))
    
    return(plot_table)
  })

```


## The renderPlotly function {.bigfont}

This code creates the plot...

```{r renderplotly, echo = TRUE, eval = FALSE}

output$plot <- renderPlotly({
  plot_ly(y = plot_table()$Var1, x = plot_table()$Freq, type = 'bar',
          orientation = 'h') %>%
      layout(margin = list(l = 150))
  })

```



# Additional notes

## Updating input options {.bigfont}

- Our app follows the following logic: <br><font size = '120%'><b>Inputs => Outputs</b></font>

- But in some cases the situation is a bit more difficult: <br><font size = '120%'><b>Inputs => Outputs => Inputs</b></font>

- Example: Data are loaded to the app via `fileInput` and the select box needs to select from variables present in the data

- There are `update*` functions in the `shiny` library for this situation - example [here](https://shiny.rstudio.com/gallery/update-input-demo.html)


## Using buttons {.bigfont}

- Sometimes it is not appropriate to re-run the computations with every change of inputs

- **Action buttons** can be used for this purpose

- The handling of action buttons in the `server` function is a bit more difficult - more details [here](http://shiny.rstudio.com/articles/action-buttons.html)

# In a nutshell

## How to build a shiny app {.biggerfont}

1. Start in the RStudio menu (File -> New File -> Shiny web app)
2. Select the layout structure of the UI
3. Arrange the widgets and outputs from Shiny gallery to the layout
4. Set how the data are loaded to the app
5. Set the reactive environment (how outputs are created from the inputs)
6. Debug the app and improve the UI
7. Deploy the app

# Questions? Comments? {.bigfont}

- The presentation available [online](https://vcepelak.shinyapps.io/shinytraining) or in [HTML file](https://raw.githubusercontent.com/vaclavcepelak/shinytraining/master/presentation_html.html)
- The shiny app available [here](https://vcepelak.shinyapps.io/MyFirstApp/)